---
title:        "`r DODschools::noTouch('metadata.yml')$document$title`"
designator:   "`r DODschools::noTouch('metadata.yml')$document$designator`"
doctype:      "`r DODschools::noTouch('metadata.yml')$document$type`"
pages:        "`r DODschools::noTouch('metadata.yml')$document$pages`"
abstract:     "`r DODschools::noTouch('metadata.yml')$abstract`"
dedication:   "`r DODschools::noTouch('metadata.yml')$dedication`"
acknowledge:  "`r DODschools::noTouch('metadata.yml')$acknowledgement`"
vita:         "`r DODschools::noTouch('metadata.yml')$vita`"
degree:       "`r DODschools::noTouch('metadata.yml')$degree`"
program:      "`r DODschools::noTouch('metadata.yml')$program`"
distro1:      "`r DODschools::noTouch('metadata.yml')$distro_thesis[1]`"
distro2:      "`r DODschools::noTouch('metadata.yml')$distro_thesis[2]`"
author:
  name:       "`r DODschools::noTouch('metadata.yml')$author$fullname`"
  dept:       "`r DODschools::noTouch('metadata.yml')$author$dept`"
  rank:       "`r DODschools::noTouch('metadata.yml')$author$rank`"
  service:    "`r DODschools::noTouch('metadata.yml')$author$service`"
  prevdegree: "`r DODschools::noTouch('metadata.yml')$author$currentDegree`"
  email:      "`r DODschools::noTouch('metadata.yml')$author$email`"
advisor:
  name:       "`r DODschools::noTouch('metadata.yml')$advisor$name`"
  department: "`r DODschools::noTouch('metadata.yml')$advisor$department`"
  rank:       "`r DODschools::noTouch('metadata.yml')$advisor$rank`"
  service:    "`r DODschools::noTouch('metadata.yml')$advisor$service`"
  prevdegree: "`r DODschools::noTouch('metadata.yml')$advisor$currentDegree`"
  phone:      "`r DODschools::noTouch('metadata.yml')$advisor$phone`"
  email:      "`r DODschools::noTouch('metadata.yml')$advisor$email`"
reader1:
  name:       "`r DODschools::noTouch('metadata.yml')$reader1$name`"
  department: "`r DODschools::noTouch('metadata.yml')$reader1$dept`"
  rank:       "`r DODschools::noTouch('metadata.yml')$reader1$rank`"
  service:    "`r DODschools::noTouch('metadata.yml')$reader1$service`"
  prevdegree: "`r DODschools::noTouch('metadata.yml')$reader1$currentDegree`"
reader2:
  name:       "`r DODschools::noTouch('metadata.yml')$reader2$name`"
  department: "`r DODschools::noTouch('metadata.yml')$reader2$dept`"
  rank:       "`r DODschools::noTouch('metadata.yml')$reader2$rank`"
  service:    "`r DODschools::noTouch('metadata.yml')$reader2$service`"
  prevdegree: "`r DODschools::noTouch('metadata.yml')$reader2$currentDegree`"
reader3:
  name:       "`r DODschools::noTouch('metadata.yml')$reader3$name`"
  department: "`r DODschools::noTouch('metadata.yml')$reader3$dept`"
  rank:       "`r DODschools::noTouch('metadata.yml')$reader3$rank`"
  service:    "`r DODschools::noTouch('metadata.yml')$reader3$service`"
  prevdegree: "`r DODschools::noTouch('metadata.yml')$reader3$currentDegree`"
sf298name:    "`r DODschools::noTouch('metadata.yml')$author$sf298name`"
contractnum:  "`r DODschools::noTouch('metadata.yml')$sf298$contractnum`"
grantnum:     "`r DODschools::noTouch('metadata.yml')$sf298$grantnum`"
prognum:      "`r DODschools::noTouch('metadata.yml')$sf298$programnum`"
projnum:      "`r DODschools::noTouch('metadata.yml')$sf298$projectnum`"
tasknum:      "`r DODschools::noTouch('metadata.yml')$sf298$tasknum`"
worknum:      "`r DODschools::noTouch('metadata.yml')$sf298$workunitnum`"
keywords:     "`r DODschools::noTouch('metadata.yml')$sf298$keywords`"
sponsor:
  title:    "`r DODschools::noTouch('metadata.yml')$sf298$sponsor$title`"
  subtitle: "`r DODschools::noTouch('metadata.yml')$sf298$sponsor$subtitle`"
  address1: "`r DODschools::noTouch('metadata.yml')$sf298$sponsor$address1`"
  address2: "`r DODschools::noTouch('metadata.yml')$sf298$sponsor$address2`"
  phone:    "`r DODschools::noTouch('metadata.yml')$sf298$sponsor$phone`"
  email1:   "`r DODschools::noTouch('metadata.yml')$sf298$sponsor$email1`"
  email2:   "`r DODschools::noTouch('metadata.yml')$sf298$sponsor$email2`"
  acronym:  "`r DODschools::noTouch('metadata.yml')$sf298$sponsor$acronym`"
  rptnum:   "`r DODschools::noTouch('metadata.yml')$sf298$sponsor$report_number`"
graddate:   "`r DODschools::noTouch('metadata.yml')$grad_date`"
date:       "`r format(Sys.Date(), '%B %Y')`"
sf298_date: "`r format(Sys.Date(), '%d-%m-%Y')`"
dissertation: "`r DODschools::noTouch('metadata.yml')$dissertation`"
cite_style: "`r DODschools::noTouch('metadata.yml')$cite_style`"
cite_shape: "`r DODschools::noTouch('metadata.yml')$cite_shape`"
num_cite: "`r DODschools::noTouch('metadata.yml')$num_cite`"
output: 
  DODschools::afit_thesis:
    highlight: tango
    includes:
      in_header:    scripts/tex/in_header.tex
      before_body:  scripts/tex/before_body.tex
      after_body:   scripts/tex/after_body.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, show = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}

#check for req'd packages, will be installed automatically if not present
list.of.packages <- c("tidyverse", "lubridate", "sas7bdat", "fpp2", "reshape2",
                      "stargazer", "knitcitations", "RefManageR", "xtable", 
                      "kableExtra", "zoo")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages, repos="http://cran.us.r-project.org")


# source file containing functions created for this analysis
source("~/Documents/Grad School/Thesis/github/afit.thesis/custom-functions.R")

#analysis required libraries
library(sas7bdat)
library(fpp2)
library(zoo)
library(lubridate)
library(reshape2)
library(knitr)
library(gridExtra)
library(tidyverse)
library(kableExtra)

#template req'd libraries
library(knitcitations)
library(RefManageR)
library(xtable)
source('scripts/R/setup.R')

BIB <- ReadBib('references/my_bib.bib')
knitcitations <- BIB[title = 'knitcitations']
refmanager    <- BIB[title = 'RefManageR']
pressure      <- BIB[key = 'randolph2016']
cite_options(citation_format = 'pandoc')

```

# Introduction

## Background

As with any large organization, the personnel management functions of the components of the Department of Defense (DoD) are concerned with personnel retention. However, since the DoD must grow all its leaders, retention is for more important and challenging.

The DoD has long offered an all or nothing 20-year retirement: stay to 20 years and you are eligible for retirement benefits, leave before 20 years and you have nothing. This 20-year goal has certainly been a positive retention motivator.

The new blended retirement system will change the all of nothing aspect of military retirement. Personnel can now lave befire 20 years with some level of retirement benefits. These new options will surely change military retention patterns.

Part of the military strategy to keep retention at desired levels is to increase pay levels of targeted personnel groups with retention bonuses. Clearly, military members offered such a bonus must consider the bonus and retaining versus civilian pay potential if the member separates.

This research is a study of military retention as affected by economic measures used as indicators of civilian employment potential.

## Scope

For both releasability and compatibility reasons, the personnel data has been aggregated to the national level, limiting the detail to which relationships can be explored. This was done to match the national economic data available, and to protect personal information of the individuals included in the analysis. 

The military personnel data concerns those serving during the 2004-2017 time frame, and the economic data matches. Some extraordinary events occurred during that period, notably the Great Recession beginning in 2008, which may have altered normal behavior and affected relationships. The U.S. military is also transitioning to a new retirement system. It is possible that any relationships revealed in this thesis will be affected by the new retirement system. 

## Assumptions and Limitations

As with any analytic endeavor, several assumptions are made in order to faciliate the modeling of real world phenomena. Perhaps most central to this thesis is the assumption that there exists at least one economic indicator (but ideally many) that helps inform an individual military member's decision to stay or leave active duty service. It is also assumed that if these variables do not directly inform retention decisions, they serve as adequate proxies for unobservable or abstract factors that do influence the individual's decision. For instance, members may not follow the movements of the Consumer Price Index (CPI), but that movement should provide information on the cost of living which may affect the decision to stay in the military. We also assume that the skills held by the Air Force officer corps are largely transferrable to civilian labor markets. This relationship may not hold equally across all Air Force career fields, however, and is investigated later in Chapter III. Standard assumptions for regression modeling and forecasting are made (independent, normal, and homoscedastic errors) and are tested, as well. 

## Outline

This chapter introduced the retention problem investigated and discussed the foundational motivations and thoughts underpinning the thesis. The next chapter reviews the related literature in detail - the efforts used to better frame the problem and previous attempts to model it. The third chapter focuses on the methodology. It documents how and why the data was attained (i.e. sources and selection criteria), as well as any transformations necessary to conduct analysis. The chapter continues by discussing the modeling procedure in detail, including general steps and specific mathmatical formulations. Lastly, the results are examined and insights or conclusions are highlighted.

\newpage

# Literature Review

## Chapter Overview

Managing personnel and modeling retention behaviors have, appropriately, long been a concern of the Department of Defense. This chapter summarizes the retention problem, examines previous research endeavors, and finally discusses the impetus for the econometric approach used in this research. 

## The Retention Problem

All organizations have some problem associated with retaining their people. This is especially true of the the military, wherein members are routinely confronted with deployments, long duty hours, and frequent relocations - factors generally not found in non-military organizations. These factors produce high stress on the military members and their families, who play a significant role in a member's retention decision \cite{fugita-lakhani-1991}. Evidence suggest that individuals serving in the military are generally more tolerant of these conflicts \cite{capon-etal-2007}, but the causes of attrition involve more than just familial concerns. Kane \cite{kane-2012} argues the military suffers from a chronic personnel mismanagement problem: Members' merit is not rewarded nearly as well as it is in the private sector, in terms of personal recognition and upward movement, partly due to heavy bureaucratic restrictions. This disparity can lead to frustration and job dissatisfaction, damaging the member's commitment to the organization and incentivizing attrition \cite{capon-etal-2007}. 

Compounding the internal frustrations, civilian labor markets can offer intense incentives for leaving. Barrows \cite{barrows-1993} details the mechanisms underpinning U.S. Air Force pilot attrition to civilian airlines, framing the problem with human capital theory. The military offers a unique opportunity for developing highly desired skill sets, placing members in positions of high stress and responsibility at early stages of professional development \cite{kane-2012}. Furthermore, evidence suggests that military as an insitution is quite adept at attracting intelligent and capable individuals \cite{asch-hosek-2004}. Providing innately talented individuals with a high degree of general and specific training fosters the development of high-performers with desirable and broadly applicable skill sets. Therein lies the problem, civilian firms are typically more flexible in their ability to compensate individuals through organizational advancement and wage, often outcompeting the military \cite{kane-2012}. These phenomena are in direct contradiction to the principles for successful retention laid out by Asch \cite{asch-1993}. Asch explains that in order for military compensation to be attractive, it needs to be at least as great as the members' the expected wages and benefits offered by civilian labor markets. Compensation should also be contigent upon performance, reflecting the individual's value to the organization, to maintain motivation and disincentivize attrition \cite{asch-1993}. In order to best determine compensation, then, it behooves the military to develop methods for anticipating the effects of labor market conditions on military members' retention decisions. 

<!--
Additionally, Carrell [6] notes that military members are sensitive to local wage disparities between the government and private sector. Intuitively this makes sense, as members receive more exposure to the wage advantages and economic conditions in their local communities than national averages. 
-->
## Previous Research

There have been many forays into personnel retention modeling and forecasting. Saving et al. \cite{saving-etal-1985} find a significant interaction between labor markets and military retention by analyzing individual career fields within the U.S. Air Force. Their results indicate that demographic factors such race and education level are influential to retention at early stages, but exhibit diminished effects as careers progress. Additionally, their work supports the conjecture that civilian wages, unemployment rates, and other economic variables affect retention. 

In 1987, Grimes \cite{grimes-1987} investigated the retention problem by applying a variety of regression methods (ordinary multiple linear regression, with logarithmic tranformations on response and/or explanatory variables) to predict officer loss estimates 6-12 months in the future. He was unable to provide adequate effects estimates or reliable predictions, concluding that the chronological nature of the data led to serial correlation errors.

Fugita and Lakhani \cite{fugita-lakhani-1991} use survey and demographic data compiled by the Defense Manpower Data Center to estimate hierarchical regression equations to describe rentention behaviors in Reservists and Guard members. Hierarchical regression models are useful when there exists some causal ordering among predictors, as is often the case with demographic and economic data. This causal relationship can lead to high multicollinearity, increasing the estimated standard error of coefficient estimates and resulting in non-significant predictors. They find that, for both officers and enlisted, retention probabilities tend to rise with increased earnings, years of service, and spousal attitude towards retention. Their work reinforces the importance of including demographic variables in retention modeling, and that wages are in the forefront of a member's mind when deciding to stay.

Gass \cite{gass-1991} takes a more general view by modeling the manpower problem in three different ways: as a Markov chain with fixed transition rates between nodes, as a minimum-cost network flow problem, and as a goal-programming problem. While potentially easier to interpret, these models can present a too-sanitized picture of an enormously complex system, particularly the current military personnel system. 

Barrows \cite{barrows-1993} analyzes retention, specifically for Air Force pilots, through the lens of human capital and internal labor market theories. He argues two points important to this thesis: the degree of specific training is inversely correlated with attrition, and that the Air Force personnel system suffers from the inefficiences typical of an internal labor market. 

To the first point, the military offers a high degree of general and specific training. General training is conducive to attrition, as it allows the individual to more easily transfer between jobs. Specific training, on the other hand, decreases worker transferability and helps improve retention. This effect is seen in differing retention rates between general pilots (e.g. cargo, heavies) and those with more specific skill sets (e.g. helicopters, fighters). One can imagine this would also reveal itself in the non-rated officer population; that is, career fields with transferable skill sets suffer more from attrition than those with specific skill sets, e.g. think logistics (general) versus aircraft maintenance (specific). 

Regarding the second point, workers are somewhat insulated from the competition posed by outside labor markets (e.g. Majors do not have to worry about civilians being hired specifically to replace them), and are paid according to position as opposed to productivity. Shielding employees from outside competition can possibly remove incentive for performance; individuals who feel more secure in their jobs may not try as hard. Not paying according to performance can be damaging in two ways: high-performers can feel undervalued and motived to leave, and under-performers could be receiving more than they produce.

Looking to the Navy, specifically Junior Surface Warfare Officers (SWOs), Gjurich \cite{gjurich-1999} found that one of the most important factors affecting retention was marriage. Single officers are more likely to leave than those with families. This actually may be a proxy for risk aversion. Those officers with dependents may be less likely to risk unemployment by leaving the military, choosing instead to keep a relatively secure job. Again, the importance of demographic factors was reinforced, but litte is said of the economic considerations.

In 2002, Demirel \cite{demirel-2002} used logit regression to analyze retention behaviors for officers at the end of their initial service obligation and at ten years of service. While the focus of this endeavor was to identify any changes in retention related to commissioning source, several other demographic factors - such as marital status, education level, and sex - were found to be statistically significant. This reinforces conclusions about demographic factors drawn by previous research efforts, and shows evidence that these trends generally apply to the military population, instead of particular service branches.

Ramlall \cite{ramlall-2004} takes a less technical approach and surveys the existing employee motivation theories to offer an explanation of how employee motivations affect retention, and how the disregard for the principles contained therein motivate attrition. Many causes are discussed, and a few are consistent (or at least common) amongst the spectrum of motivation theories. When wages and promotions are not seen as tied to performance, individuals are disincentivized and do not feel as loyal to the institution. Also, a lack of flexibility within job scheduling and structure is seen as disloyal or disrepectful to the individual. Lastly, when managers fail to act as coaches or are not seen as facilitators to employees' careers, turnover rates tend to be greater. Given that civilian labor markets are generally more flexible in both pay structure and work scheduling, this research underpins the importance of incorporating civilian labor market conditions.

More recently, Schofield \cite{schofield-2015} employs a logisitic regression model to identify key demographic factors influencing the retention decisions of non-rated Air Force Officers. She finds that career field grouping, distinguished graduate status at commissioning source, years of prior enlistment, and several other structural variables were significant. She then utilizes these factors to generate a series of survival functions describing retention patterns and behavior. Again, the importance of demographic factors is reinforced. However, any possible effects of economic factors were unexplored.

Looking at the rated officer corps, Franzen \cite{franzen-2017} takes a similar approach to Schofield \cite{schofield-2015} using logisitic regression to identify significant factors and generating survival functions. However, Franzen's work differs from Schofield by choosing to also assess the influence of economic, demographic, and other variables exogenous to the military. She finds that marital status, number of dependents, gender, source of commissioning, prior enlisted service, and the New Orders value from the Advance Durable Goods Report were all significant. The first couple support the aforementioned notion that familial strain caused by military service affects retention, the next few (gender, source of commissioning, and prior service) reaffirm the work conducted by Schofield. The last variable, New Orders, suggests that indicators of economic health play some role in retention decisions. This last observation is a motivation for this thesis.

In that vein is the work conducted by Jantscher \cite{jantscher-2016} where she conducts correlation analysis to determine the relationship between a host of economic indicators and retention rates for each Air Force Specialty Code (AFSC). The results of the preliminary correlation analysis provide a subset of economic indicators shown to be significantly correlated with retention, such as unemployment rates, gross national savings, real GDP growth, etc. She then attempts to form a regression model to forecast retention, but was unable due to build a model due high multicollinearity between many of the indicators. Nonetheless, her correlation analysis provides a starting point from which additional modeling techniques may be applied.

## Insights

Several key themes arise based on the review of the literature:

- Demographic and economic factors can play a significant role in a member's attitude towards retention;
- Military members are aware of and incorporate opportunities in the civilian labor market when deciding to remain in or leave military service;
- Logistic regression on demographic data yields promising results when predicting whether an individual will remain in service, but may be innappropriate for modeling aggregate trends; and
- Effects estimation of economic factors through regression can be difficult, as many indicators are highly correlated.

What is also apparent is that there are several topics yet unexplored:

- Modeling the military population with performanced-based pay structures and advancement schemes to examine effects on retention;
- Exactly how comparable the military population is to the civilian, how easily the professional skills sets exhibited by the former transfer to the latter; and
- Applying other forecasting techniques (ARIMA, Exponential Smoothing, Dynamic Regression) to the retention problem.

This thesis focuses on the last point. The goal is to forecast Air Force Non-rated officer retention with a dynamic regression model in order to estimate the effects of different economic indicators. This is covered in the next chapter.


<!-- Reviewing the existing literature, several themes are apparent: demographic and economic factors play a large role in a member's decision to leave the military - skill transferability, spousal opinion, performance rewards, etc. all affect retention. This is neither new nor novel information. However, most recent technical analysis work has been focused on the demographic aspects. Though much theory points towards the influence of the economic environment, little technical analysis has focused on more than just unemployment or civilian wages. It is the aim of this thesis to expand, if only marginally, that consideration. -->



\newpage

# Analysis and Results

```{r}
# set directory for lazy data referencing
setwd("~/Documents/Grad School/Thesis/github/afit.thesis/Data")

# Here we read in all our data and combine the relevant sets for something workable

# Personnel Data
afsc_list <- read.sas7bdat("lu_ao_afs.sas7bdat")
assigned <- read.sas7bdat("assigned_levels.sas7bdat") %>% 
  spread(AFS, Assigned)
attrition <- read.sas7bdat("separations_count.sas7bdat") %>% 
  spread(AFS, Separation_Count) 

# The dates in the personnel data are in total days since 1 Jan 1960 (SAS default)
# We'll need to reformat the date into something readable and mergeable with our 
# econ set. Also, we create a new column - the total number of separations across 
# all AFSCs. We'll also create totals for different categories of officers: rated, 
# non-rated, line, etc.
attrition <- mutate(attrition, Total = rowSums(attrition[-1], na.rm = TRUE)) %>% 
  mutate(temp = EOP_Date * 86400) %>% 
  mutate(Date = as.POSIXct(temp, origin = "1960-01-01")) %>% 
  within(rm(temp, EOP_Date))

# repeat prodcedure for assigned data
assigned <- mutate(assigned, Total = rowSums(assigned[-1], na.rm = TRUE)) %>% 
  mutate(temp = EOP_Date * 86400) %>% 
  mutate(Date = as.POSIXct(temp, origin = "1960-01-01")) %>% 
  within(rm(temp, EOP_Date))

# Econ Data
econ_data <- list.files(pattern = "*_natl.csv") %>% 
  # Import data sets
  lapply(read_csv) %>% 
  # Merge all sets 
  Reduce(function(x,y) merge(x, y, by = "DATE"), .) %>% 
  # Store data as tibble
  as.tibble() 

# Now, because gdp per cap is observed only on a quarterly basis, we add it separately
# in order to handle the missing values resulting from a merge. This merge is 
# from the previous in that it keeps all values from the existing data set, creating NAs
# where gdp_per_cap does not match with other observations. Merging the quarterly 
# gdp_per_cap data with the monthly indicators results in NAs in gdp_per_cap where 
# the dates do not match. We then handle NAs by extending the quarterly records throughout 
# their respective quarters (e.g. the GDP per cap for Q1 2014 is applied to Jan-Mar 2014).
# Simultaneously, we will rename the variables for interpretability.

# Read in GDP per capita
gdp_per_cap <- read_csv("real09_gdp_percap.csv")
# Combine gdp per cap with econ_data, using a left-join (all.x = TRUE) to preserve the main data set
econ_data <- merge(econ_data, gdp_per_cap, by = "DATE", all.x = TRUE) %>%
  as.tibble() %>%
  # Rename column headers to something more meaningful
  select(Unemployment.Rate.Adj = UNRATE, Unemployment.Rate.NonAdj = UNRATENSA,
         CPI.Adj = CPIAUCSL, Nonfarm.Jobs.Adj = JTSJOL, 
         Nonfarm.Jobs.NonAdj = JTUJOL, Labor.Force.Participation = LNS11327662, 
         Labor.Market.Momentum = FRBKCLMCIM, Real.GDP.Per.Capita = A939RX0Q048SBEA, 
         Nonfarm.Quits.Adj = JTSQUL, Date = DATE)

# The na.locf() command below carries a value forward through NAs until the next non-empty value is met; 
# this is how we choose to represent the gdp per capita through an entire quarter
econ_data$Real.GDP.Per.Capita <- as.numeric(econ_data$Real.GDP.Per.Capita) %>% 
  na.locf()

# However, the date variables differ slightly between the econ and personnel sets:
# Though both represent monthly observations, the econ set default to the first 
# of each month, and the personnel data defaulted to the last 
# (e.g. October 2004 is represented as 01-10-2004 in the former, and 31-10-2004 in the latter)
# To handle this, we'll create new date variables in each set that have the days 
# trimmed off, then merge. Merging isn't strictly necessary, but it is a convenient
# way to only keep those observations common to both data sets.
econ_data <- mutate(econ_data, "Date1" = paste(year(Date), month(Date)))
attrition <- mutate(attrition, "Date1" = paste(year(Date), month(Date)))

# Merge data
df <- merge(econ_data, attrition, by = "Date1")

# Next, we see many NAs within the attrition data set. Given the data's nature, 
# our intuition was that these missing values aren't a result of encoding error 
# or similar, but rather an indication that no separations occurred during 
# that period (i.e. NAs indicate no separations were observed, instead of
# indicating some sort of error). This intuition was confirmed by the data's
# provider, HAF/A1FPX.
df[is.na(df)] <- 0

# Next we'll go ahead and drop all of our date variables. When we use df
# to create a time series object, the date variables become redundant.
df <- df[, !(names(df) %in% c("Date1", "Date.x", "Date.y"))]

# Now we'll initialize the time series object - start = Oct 2004, freq = 12 - 
# and create the validation and training sets. Since we're only really interested
# in the Total column for modeling purposes
df.ts.1 <- ts(df, start = c(2004, 10), frequency = 12)
train.ts.1 <- subset(df.ts.1, end = 127)
val.ts.1 <- subset(df.ts.1, start = 128)


```


## Data Composition

### Introduction

Before any predictive or descriptive analysis can begin, the data must be understood. Every data set has its idiosyncracies, its own unqiue challenges. Understanding these characteristics and the meaning of the data - what the variables represent and how they might interact with each other - is key to any successful analytic endeavor. Here, the data are described in detail - its sources, meaning, and peculiarities.

### HAF/A1XDX

The Strategic Analysis branch of the Force Management Division (AF/A1XDX) provided the data on Air Force personnel used in this thesis. The data are extracted from the Military Personnel Data System (MilPDS), a database containing Air Force personnel data for every airman over his or her career. The data are input by trained personnelists or are automatically updated within the system (e.g., age will automatically increase). The data were originally split into two separate `.sas7bdat` files, one containing monthly attrition numbers for each Air Force Specialty Code (AFSC) and the other detailing monthly assigned levels for each AFSC. Each file contains information starting in October of 2004 through September of 2017, for a total of 156 observations across 67 AFSCs. 

### Federal Reserve Bank of St. Louis

The Federal Reserve Bank of St. Louis is one of 13 entities which comprise the United States' central bank (the others being 11 regional reserve banks and the Board of Governors). As a whole, the central bank is responsible for deciding on and enacting monetary policy for the U.S. They maintain expansive databases containing information about the economic environment - financial data, national employment statistics, private sector business data, etc. Fortunately, the Federal Reserve Bank of St. Louis offers public access to the Federal Reserve Economic Data (FRED) database via online interface. From here, historical data on several economic indicators were retrieved: the nation unemployment rate (both seasonally adjusted and non-adjusted), the labor force participation rate (LFPR), job openings (adjusted and not), total nonfarm job quits, the FED's labor market momentum index, real GDP per capita, and the consumer price index (CPI). Each indicator consists of monthly recordings across varying time spans (e.g. 1990-2016 vs 2001-2017).

The LFPR is the percentage of the population actively employed or looking for employment. Its movement can give insight into the strength of the economy - e.g. rising participation is usually associated with growth. When paired with unemployment rates, the LFPR can also reveal people's attitude about the economy. For example, the steady decline of participation from 2010 onward (seen in Figure \@ref(fig:lfpr-unemployment)) might indicate that the decrease in unemployment over the same period is somewhat exaggerated; people seeking, but unable to find, work may become discouraged and exit the labor force, artificially decreasing the unemployment rate. It is possible that this perception of economic health affects retention decisions. In this thesis, LFPR is restricted to members of the civilian labor force with at least a baccalaureate and no younger than 25 years of age, the subset most closely related to officers.

```{r lfpr-unemployment, fig.align="center", fig.cap="Participation and Unemployment", fig.width=6, fig.pos="H", fig.height=4}

# subset LFPR and unemployment, then graph to reinforce a point made above

# data.frame("LFPR" = combined_1[,"Labor.Force.Participation"],
#            "Unemployment Rate" = combined_1[,"Unemployment.Rate.Adj"]) %>% 
df.ts.1[,c("Labor.Force.Participation", "Unemployment.Rate.Adj")] %>%   
  autoplot(facets = TRUE) + 
  geom_smooth(size = .5, alpha = 0) + 
  scale_x_continuous(limits = c(2005, 2016), name = "Year") +
  scale_y_continuous(name = "Percent of Population")

```


It is assumed that the skillsets of our target population (Air Force officers) are most transferrable to those jobs covered by nonfarm payrolls. Nonfarm is a category that excludes proprietors, private household employees, unincorporated self-employment, unpaid volunteers, and farm employees \cite{fred-nonfarm-2017}. Job quits are generally voluntary separations and may reflect workers' willingness to leave; it may be that the a higher propensity to volutarily leave a job translates to a positive outlook on obtaining another. 

The labor market momentum index compares current labor market conditions to historical averages. A negative value indicates conditions below the long-term average, and a positive value indicates favorable conditions. The CPI examines the weighted average price of a basket of consumer goods and services; it is used to estimate the cost of living. There is some uncertainty involving employment in separation from the military, so cost of living information may be especially important to that decision as the military is excluded from CPI statistics.

By including these variables in a regression model and estimating their effects, this work seeks to capture military members' perceptions of economic health and job prospects, and use that information as a means to forecast attrition.

```{r econ-var, fig.cap='Selected Economic Indicators'}

panderOptions('table.alignment.default',
   function(df) ifelse(sapply(df, is.numeric), 'right', 'left'))

vars <- data.frame(Variable = c('Labor Market Momentum Index', 'CPI', 
                              'Nonfarm Jobs Openings', 'Real GDP per Capita', 
                              'Nonfarm Job Quits', 'Unemployment Rate',
                              'Labor Force Participation Rate'),
                  Description = c('Compares current market conditions to long-run average', 
                                 'Weighted average price of a basket of goods and services', 
                                 'Unfilled positions at the end of the month in the nonfarm sector', 
                                 'Measure of economic output per person, adjusted for inflation', 
                                 'Voluntary separations from jobs in the nonfarm sector', 
                                 'Percentage of unemployed individuals in the labor force', 
                                 'Percentage of the population either employed or actively seeking work'))


kable(vars, format = "latex", booktabs = TRUE, caption = "Selected Economic Indicators") %>% 
  kable_styling(full_width = FALSE, latex_options = "scale_down") %>% 
  row_spec(0, bold = TRUE)


# pander(vars, split.cell = 70, split.table = Inf)

```


### Cleaning and Preparation

Perfect data is rarely found or received outside of the classroom, and such is the case here. Before exploration and modeling, several steps are taken to produce a useable data set.

The personnel data is first converted from long to wide format. Originally, the personnel data comes with three variables: AFSC, Date, and Separations. This form is not conducive for modeling, so a new variable is created for each category in AFSC containing the associated separation counts. This procedure generates missing values, and they need to be dealt with appropriately. Missing values can result from several underlying issues: data storage corruption, entry errors, miscommunication between software, etc. Luckily, none of those apply here. Since the attrition data is a monthly count of people exiting USAF service, the intuition is that these missing values simply represent a lack of an observation (i.e. zero separations). This is confirmed by the data's provider, and so all missing values are replaced with zero.
Initially, observation dates are stored as the number of days since 1 Jan 1960 (the standard for SAS). This is transformed into YYYY-MM-DD, in order to merge with the economic data. Several columns are added, each the total separations for a specific subset of the officer corps: rated, non-rated line, non-line, etc. These sums will be the responses used for modeling.

The economic data do not require much treatment, as they come from a professionally managed database. One of the indicators, real GDP per capita, occurs in quarterly intervals - the rest are monthly. To make data comparable, the quarterly values are applied across each month in the quarter (e.g. the observation for Q1 2006 is applied over January, Febuary, and March 2006). Then, variables are also renamed for clarity. Finally, economic data are merged with the personnel through an inner join, preserving only those observations with dates common to both data sets. 


## Model Selection

### Introduction

This section documents the modeling process, the analytic decisions at each juncture and their reasons. Before starting, however, several key concepts and definitions must be reviewed. 

General modeling practices involve horizontally splitting the original data set into at least two, sometimes three, subsets. This is done to ensure model fitting and assessment are independent processes. There are many ways to generate these subsets, each particular to the structure of the data. With time-series data, as in this thesis, the typical approach is to retain roughly the first 80 percent for model fitting, leaving the rest for model assessment. These two sections are respectively known as the training and validation sets. The training set is used to estimate model parameters, which are then used for predictions on subsequent observations. These predictions are compared against the validation set - actual, observed data - as a means of assessing model performance. Once a final model has been designated, the training and validation data are re-combined and used to produce out-of-sample forecasts. The training/validation approach is applied to each modeling technique discussed in this thesis.

This endeavor utilizes two models for forecasting: naive and dynamic regression. The former is a far simpler technique and is only used as a baseline. The latter is a bit more complex. Dynamic regression has two major components, regression and time-series, each with their own assumptions and requirements. Regression models with multiple predictor variables assume independence of those predictors (also called regressors or exogeneous variables). All regression models assume errors are normally and independently distributed around zero with constant variance. <!-- NEEDS MORE -->


### Initial Exploration

The first step is to examine the data visually. Plotting one response, total separations over all career fields, in Figure \@ref(fig:response-plot) shows significant spikes during 2005, '06, '07, and '14. We know that during these periods, special spearation incentive programs were introduced in order to artificially downsize the air force. The effects of these periods merit investigation later on, as they could negatively affect model performance. 
```{r response-plot, fig.align="center", fig.cap="Monthly Officer Separations", fig.pos="H", fig.width=5, fig.height=3.5}
autoplot(df.ts.1[,'Total'], ylab = "Total Separations")
```

No seasonality is immediately obvious in Figure \@ref(fig:response-plot). However, if each year is plotted separately, a clearer picture emerges. First, Figure \@ref(fig:response-season-plot) shows that the extreme points noticed noticed above seem to be relegated to the November-December time frame. Second, it is easier to witness the seasonality: bowing across the year, with higher counts at the beginning and end.

```{r response-season-plot, fig.align="center", fig.cap="Seasonal Plot: Total Separations", fig.pos="H", fig.width=5, fig.height=3.5}
p <- ggseasonplot(df.ts.1[,'Total'], year.labels = TRUE, year.labels.left = TRUE) +
  ylab("Total Separations") + 
  ggtitle("") + 
  theme(legend.position = "none")
p  

```

Considering these plots, we expect that a seasonal model performs best and that we will have to perform some alteration to accomodate the outiers. To confirm, we fit naive models to the data and examine to results. Beyond revealing seasonality and outlier effects, fitting naive models establishes a baseline to compare against later models. Naive models are very simplistic, so if later models perform worse or only marginally better, it implies they are not capturing much information.

```{r}
n.1 <- naive(train.ts.1[,"Total"], h = dim(val.ts.1)[1])
sn.1 <- snaive(train.ts.1[,"Total"], h = dim(val.ts.1)[1])

n.1.error <- accuracy(n.1, val.ts.1[,"Total"])
sn.1.error <- accuracy(sn.1, val.ts.1[,"Total"])
```

Figure \@ref(fig:n-sn-forecast) gives evidence to the negative effects of outliers, notice the large confidence intervals surrounding the naive forecast and the 2014 spike carried through in the seasonal forecast.

```{r n-sn-forecast, fig.align="center", fig.pos="H", fig.cap="Simple and Seasonal Naive Forecasts", fig.width=5}
p <- autoplot(n.1) +
  autolayer(val.ts.1[,"Total"]) +
  theme(legend.position = "none") +
  ylab("") + 
  xlab("")

q <- autoplot(sn.1) +
  autolayer(val.ts.1[,"Total"]) +
  theme(legend.position = "none") +
  ylab("") +
  xlab("")

grid.arrange(p, q, nrow=2, left = "Total Attrition", bottom = "Time")

```

Tables \@ref(tab:n-err) and \@ref(tab:sn-err) show different error metrics for each of the two models. By mean square error (MSE), the seasonal model generally fits the training data better, possibily indicating presence of seasonality effects. However, there is a large disparity between validation MSEs, possibly caused by the major spike in 2014 - reaffirming the earlier intuition about outlier effects.

```{r n-err, fig.align="center", fig.pos="H", fig.width=4, fig.cap="Naive Results"}
kable(n.1.error[,1:6], caption = "Naive Results", digits = 3, align = 'c')
```
```{r sn-err, fig.align="center", fig.pos="H", fig.width=4, fig.cap="Seasonal Naive Results"}
kable(sn.1.error[,1:6], caption = 'Seasonal Naive Results', digits = 3, align = 'c')
```

It is known that during years 2005, '06, '07, and '14 special separation programs were implemented. Given the effect those years appear to have on modeling, they must be accomodated before continuing. Before deciding how, the explicit points in question need to be identified. To help, refer back to Figure \@ref(fig:response-season-plot). We note above that the spikes generally occur in November and December. However, the observations from 2005 are close enough to those from other years that they may have resulted naturally. We do not want to remove too much information from the data set, only that which is misleading. So, November and December observations from 2006, '07, and '14 are selected for replacement. 

Given the seasonality in the data set, the replaced values should stem from matching observations in previous years, as opposed to previous observations within the same year. The outliers are replaced with the arithmetic mean of all years not being replaced (e.g. November 2006, '07, '14 is replaced with the mean separations in November for all other years).

```{r}
# To handle these, we'll calculate the average values of all other years during 
# months and replace the current values. First, let's create slices of our
# response containing the months we're concerned with - December and November.
# We also want to grab the correpsonding indices for updating our series later.

dec <- subset(df.ts.1[,'Total'], month = 12)
dec.ind <- which(cycle(df.ts.1[,'Total']) == 12)
nov <- subset(df.ts.1[,'Total'], month = 11)
nov.ind <- which(cycle(df.ts.1[,'Total']) == 11)

# oct <- subset(df.ts.1[,'Total'], month = 10)
# sep <- subset(df.ts.1[,'Total'], month = 9)

# Referring back to p, and combining the graphical insights with information
# from the data's sponsor, we assume that 2006, '07, and '14 are the years which
# saw the largest effects from the separation incentive programs - i.e. artificial
# attrition. Those correspond the the 3rd, 4th, and 11th indices. So now, we 
# replace those observations with the average of the non-aberrant years.

dec[c(3,4,11)] <- mean(dec[-c(3,4,11)])
nov[c(3,4,11)] <- mean(nov[-c(3,4,11)])

# And finally, we place these values back into the original series.
df.ts.1[dec.ind, 'Total'] <- dec
df.ts.1[nov.ind, 'Total'] <- nov
```

Replotting the repsonse in Figure \@ref(fig:response-plot-2) shows a much better behaved data set. The data look fairly stationary, setting the stage for developing more complex forecasting models.

```{r response-plot-2, fig.align="center", fig.cap="Separations - Outliers Removed", fig.pos="H", fig.width=5, fig.height=3.5}
autoplot(df.ts.1[,'Total'], ylab = "Total Separations")
```

With the outliers replaced, seasonal effects are much more apparent (see Figure \@ref(fig:response-season-plot-2) below), further enforcing the need for a seasonal model.

```{r response-season-plot-2, fig.align="center", fig.cap="Seasonal Plot: Outliers Removed", fig.pos="H", fig.width=5, fig.height=3.5}
ggseasonplot(df.ts.1[,'Total'], year.labels = TRUE, year.labels.left = TRUE) +
  ylab("Total Separations") + 
  ggtitle("Seasonal Plot: Total Separations") + 
  theme(legend.position = "none")
```

```{r}
# store split index
set.split <- 127
# New train and val sets
train.ts.2 <- subset(df.ts.1[,'Total'], end = set.split)
val.ts.2 <- subset(df.ts.1[,'Total'], start = set.split+1)

# Train models and generate errors
n.2 <- naive(train.ts.2, h = length(val.ts.2))
sn.2 <- snaive(train.ts.2, h = length(val.ts.2))

n.2.error <- accuracy(n.2, val.ts.2)
sn.2.error <- accuracy(sn.2, val.ts.2)
```

Table \@ref(tab:season-rmse-compare) compares the seasonal naive RMSEs before and after imputating the identified outliers. The results indicate that removing and replacing the extreme values for November and December improved the model. This is further reflected by the forecast (shown in blue) in Figure \@ref(fig:sn-forecast-2), which follows the validation data (shown in orange) more closely than those in Figure \@ref(fig:n-sn-forecast). Overall, these results imply that imputation of the selected observations was useful.

```{r season-rmse-compare, fig.align="center", fig.pos="H", fig.cap="Seasonal Naive Comparison - Before and After Imputation"}
outlier.compare <- data_frame("Raw Data" = sn.1.error[,2],
                              "Imputed Data" = sn.2.error[,2])
row.names(outlier.compare) <- c("Training", "Validation")

kable(outlier.compare, row.names = TRUE, caption = "Seasonal Naive RMSE Comparison",
      digits = 3, align = 'c')
```

```{r sn-forecast-2, fig.align="center", fig.pos="H", fig.cap="Seasonal Naive Forecast After Imputation", fig.width=5, fig.height=3}
autoplot(sn.2) +
  autolayer(val.ts.2) +
  theme(legend.position = "none") +
  ylab("Total Attrition")
```


### Dynamic Regression

Naive models are simple, regression models adequately invlove exogenous predictor variables, and time series model adequately handle autoregressive components of data. Individually these models are useful, but cannot handle both exogenous and autoregressive components.

Dynamic regression is a multivariate regression model with an ARIMA model on the errors. The regression piece allows use of independent variables in predicting a response, and the ARIMA portion helps model the autoregressive information which can exist in time-series data. The general formulation of a dynamic regression model with ARIMA(1,1,1) errors:

$$ y_t = \beta_0 + \beta_1x_{1,t} + ... + \beta_kx_{k,t} + n_t$$
where,

$$ (1-\phi_1B)(1-B)n_t = (1+\theta_1B)e_t $$
and $e_t$ is white noise.

Fitting a dynamic regression model requires taking several steps to ensure key assumptions are not violated. First is the issue of collinearity. Collinearity between predictor variables implies a dependent relationship and can lead to innaccurate coefficient estimates, a result contrary to the goal of any modeling effort. To avoid this pitfall, a correlation matrix of possible regressors is built. A correlation matrix shows how collinear each pair of indicators is. A high correlation coefficient between indicators implies collinearity, meaning the they should not be used together in the model. Figure \@ref(fig:heat-map-1) shows the correlation between all pair-wise combinations of variables in the economic data set. There are many instances of collinearity, which is expected - many economic indicators are constructed from similar information. There are some independent subsets, though, and these will be the candidates for the dynamic regression model.

```{r heat-map-1, fig.align="center", fig.pos="H", fig.cap="Correlation Matrix - Economic Indicators", fig.width=7, fig.height=5}
df[which(names(econ_data) %ni% c("Date", "Date1"))] %>% 
  cor() %>%
  round(2) %>% 
  reorder.cormat() %>%
  get.upper.tri() %>% 
  melt(na.rm = TRUE) %>% 
  ggplot(aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Collinearity") +
  theme_minimal() + # minimal theme
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1)) +
  coord_fixed() + 
  geom_text(aes(Var2, Var1, label = value), color = "black", size = 3) +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.ticks = element_blank(),
    legend.justification = c(1, 0),
    legend.position = c(0.6, 0.7),
    legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1, 
                               title.position = "top", title.hjust = 0.5))
```

Given their low correlation, the Unemployment Rate (Adj.), Labor Force Participation Rate, and Labor Market Momentum Index are selected as independent variables. In order to ensure the assumptions made by the ARIMA piece of the model, the stationarity of the regressors is checked. Though trend and seasonality components can be incorporated through regression techniques, ARIMA models require stationarity. Non-stationary variables can produce inconsistent coefficient estimates, even if they are independent. To assess, a plot of the three indicators in Figure \@ref(fig:econ-vars-1) is generated. 

```{r econ-vars-1, fig.align="center", fig.pos="H", fig.cap="Economic Indicators - Raw", fig.height=5, fig.width=7}
# grab index of econ vars: LFPR, Unem.Adj, LMM
econ.vars <- which(names(df) %in% c("Labor.Force.Participation",
                                            "Unemployment.Rate.Adj", 
                                            "Labor.Market.Momentum"))

# Now that we've selected our regressors, we need to check for stationarity. We're
# looking for evidence of non-zero trend, seasonality, etc.

autoplot(df.ts.1[,econ.vars], facets = TRUE) +
  ylab("")
```

Each of the three indicators show evidence of a trend (i.e. are non-stationary), and to handle this the data are differenced. The resulting data are shown below in Figure \@ref(fig:econ-vars-2). 

```{r econ-vars-2, fig.align="center", fig.pos="H", fig.cap="Economic Indicators - Differenced", fig.height=5, fig.width=7}

econ.vars.d <- diff(df.ts.1[,econ.vars])

autoplot(econ.vars.d, facets = TRUE) +
  ylab("")

```

Simple differencing appears to produce the desired effect; there is no evident trend or seasonality in any indicator. It is important to note the regressors now show the month-to-month change, which will be pertinent when interpreting results. <!--do I want to talk about the stationarity of the response?--> 

```{r}
# Oh, yea - way better. Differenced, these variables look useable. Now while 
# we're looking at differencing, we should look to see if our response also
# needs to be differenced. Remember, we have this:

# autoplot(df.ts.1[,"Total"]) +
#   ylab("Total Attrition")

# Hmm, nothing too crazy, actually - is there any statistical evidence
# for differencing?

# regular differencing
#ndiffs(df.ts.1[,"Total"])
# seasonal differencing
#nsdiffs(df.ts.1[,"Total"])

# Neat! However, before we can build the model, we have to account for the
# differencing performed on our regressors. With simple differencing, we lose
# the first observation, and so we remove the first observation from our 
# response.

#head(df.ts.1[,"Total"])
#head(econ.vars.d[,1])

response.ts <- ts(df.ts.1[-1,"Total"], start = c(2004, 11), frequency = 12)

#head(response.ts)
```

With stationary and independent economic indicators, we can move on from regression to the ARIMA piece of the model. Up to six parameters must be specified: the order of autoregression, degree of differencing, and order of the moving average ($p,d,$ and $q$, respectively) and their seasonal counterparts ($P,D,$ and $Q$). For the sake of time, these are not manually determined; instead, a brute-force algorithm is used. A range is specified for each parameter, and a model is fit for every combination within the specified ranges; the model with the lowest corrected Akaike Information Criteria (AICc) is selected. For the first, and all subsequent, dynamic regression models in this work, the following ranges/values were used: 
$p,q \in [0,5]$, $d,D = 0$, and $P, Q \in [0,2]$. 


```{r, results="hide"}
# We also need to split up the econ vars into training and test sets

# store split index
set.split <- 126

# subset our response
train.ts.3 <- subset(response.ts, end = set.split)
val.ts.3 <- subset(response.ts, start = set.split+1)

# subset econ variables
econ.vars.d.train <- subset(econ.vars.d, end = set.split)
econ.vars.d.val <- subset(econ.vars.d, start = set.split+1)

# We'll utilize the auto arima function from fpp2
dyn.reg.1 <- auto.arima(train.ts.3, xreg = econ.vars.d.train, trace = TRUE,
                        stepwise = FALSE, approximation = FALSE)
```

For this first pass, the model selected was a regression model with a fourth-order moving average and first-order seasonal autoregression on the errors. More explicitly:

$$ y = \beta_0 + \beta_1x'_{1,t} + \beta_2x'_{2,t} + \beta_3x'_{3,t} + n_t$$
where, 
$$ (1-\Phi_1 B^{12})n_t = (1 + \theta_1B + \theta_2B^2 + \theta_3B^3 + \theta_4 B^4)e_t $$
and, 
$$ x'_{i,t} = x_{i,t} - x_{i,t-1}$$
The coefficient estimates are shown in Table \@ref(tab:dynreg1-coeff): 

```{r dynreg1-coeff, fig.align="center", fig.pos="H", fig.cap="Estimated Coefficients - Initial Model"}

# grab coefficients and s.e. from model
coeffs <- round(dyn.reg.1$coef, digits = 3)
stderr <- round(sqrt(diag(dyn.reg.1$var.coef)), digits = 3)

# create data table
estimates <- bind_rows(coeffs, stderr)

rownames(estimates) <- c("Coeff", "StdErr")
colnames(estimates) <- c("$\\theta_1$", "$\\theta_2$", "$\\theta_3$", 
                         "$\\theta_4$", "$\\Phi_1$", "$\\beta_0$", 
                         "$\\beta_1$", "$\\beta_2$", "$\\beta_3$")

kable(estimates, escape = FALSE, caption = "Estimated Coefficients - Initial Model",
      align = 'c', digits = 3, booktabs = TRUE, format = "latex") %>% 
  kable_styling() %>% 
  row_spec(0, bold = TRUE)
```


Model assessment involves analysis of the residuals. When inspecting the residuals, we are looking for evidence of remaining autocorrelation, satisfaction of normality assumptions ($e_t \sim N(0,\sigma^2)$), and outlier effects. Figure \@ref(fig:dynreg1-resid) provides plots used to answer those questions. The top subfigure plots the raw model residuals, and is used to identify possible trends, seasonality, or heteroscedasticity. Fortunately, none of those features are apparent. The bottom-left is used to examine significant autocorrelation in the residuals; significant correlations would indicate a possible violation of the independence of the residuals. The current model's results only show one lag-period with significant autocorrelation, which may mean that there is information unnaccounted for by the model. Overall autocorrelation, however, appears to be insignificant, as further evidenced by the results of a Ljung-Box test for autocorrelation (Table \@ref(tab:dynreg1-boxtest)). The bottom-right plot shows a histogram of the residuals, comparing the raw distribution against the ideal normal. The plot shows slight skewness, but overall the data appear normal. Thus, there is little - if any - misbehavior in the model's residuals.

```{r dynreg1-resid, fig.align="center", fig.pos="H", fig.cap="Initial Attrition Model - Residual Analysis", fig.height=5, fig.width=6}
# The first thing we'll want to check, after the model summary, is how our 
# residuals behave. Do they appear to satisfy normality assumptions? Are there 
# any outliers? Evidence of leftover autocorrelation? Essentially, we're 
# checking to see if there's anything other than white noise in the error term.

# No - to all of the above. ACF plots look clean, raw residuals seem 
# noisy, and also have a roughly normal distribution. Furthermore, a Ljung-Box
# test shows no evidence that the data aren't normal (p: 0.8121).
checkresiduals(dyn.reg.1, test = FALSE)
```

```{r dynreg1-boxtest, fig.align="center", fig.pos="H", fig.cap="Initial Model - Autocorrelation Test"}

p <- Box.test(dyn.reg.1$residuals, lag = 24, type = "Lj", fitdf = 9)

box.results <- data_frame("Test type" = p$method, 
                          "Test statistic" = round(p$statistic, digits = 3),
                          "p-value" = round(p$p.value, digits = 3))

kable(box.results, caption = "Initial Model - Autocorrelation Test")
```


Forecasts are generated from the training data and compared against the validation data.
Figure \@ref(fig:dynreg1-forecast) plots the training and validation data against the model predictions. Large movements are generally captured, even if not perfectly forecast. To compare modeling performance, the RMSEs are from our models are compared in Table \@ref(tab:model-compare-1).


```{r dynreg1-forecast, fig.align="center", fig.pos="H", fig.cap="Initial Model - Forecasts Against Validation Data", fig.height=3.75, fig.width=6.5}
# Let's generate some forecasts then
dyn.reg.1.f <- forecast(dyn.reg.1, xreg = econ.vars.d.val, h = 20)

# And then plot those forecasts over the actual data
autoplot(dyn.reg.1.f) + 
  autolayer(dyn.reg.1$fitted) +
  autolayer(val.ts.3) +
  theme(legend.position = "none") +
  ylab("Total Attrition")

```


```{r model-compare-1, fig.align="center", fig.pos="H", fig.cap="Model RMSE Comparison"}
# We'll also take a look at some accuracy measures. According to RMSE and MASE,
# the dynamic regression model performs better than a seasonal naive estimation;
# that's good news - we're getting closer towards accurate forecasting.
dyn.reg.1.error <- accuracy(dyn.reg.1.f, val.ts.3)

err.compare <- data_frame("Simple Naive" = n.2.error[,2],
           "Seasonal Naive" = sn.2.error[,2],
           "Dynamic Regression" = dyn.reg.1.error[,2])

rownames(err.compare) <- c("Training", "Validation")
  
kable(err.compare, caption = "Model RMSE Comparison", align = 'c', digits = 3)

```

Dynamic regression demonstrates greater ability to forecast the attrition data than the naive models. However, the high standard errors of the regression coefficients (Table \@ref(tab:dynreg1-coeff)) indicate that none of the economic indicators are statistically significant. This means the ARIMA model handles all the forecasting and the regression provides little insight. Essentially, the economic predictor variables do not explain much of the data variability. This could be for several reasons:

- With differencing, the indicators represent month-to-month changes. For most observations in the data set those changes are marginal, resulting in an insignificant effect on attrition - numerically. 

- As they are, the indicators show only the previous month's change. That is, the regression coefficients represent the effects last month's changes have on this month's attrition. Intuitively, this does not seem correct. Voluntary separation from the military is a long, bureaucratic process; as such, one would think the time-line is well beyond a month.

- The economic and personnel data are both aggregated to the national level. It is possible that such a degree of aggregation includes so much noise that any economic effects are being masked.

Unfortunately, there is not much that can be done about the first point. As mentioned earlier, the indicators must be stationary in order to ensure the reliability of any potential effects, and the data must be differenced to be stationary. The rest of the thesis, then, focuses on the last two points. It is simple enough to introduce lagged regressors, and that process is introduced next. As for aggregation, the economic data cannot be broken down, but the personnel data can (to an extent). Currently, the personnel data encompasses all AFSCs. It may be that different subsets of the Air Force officer corps are more sensitive to economic conditions than others, a possibility investigated in subsequent analysis.

### Lagged Indicators

### Officer Subsetting


\newpage

# Conclusions and Insights

# Appendix A. R Code {-}

```{r eval=FALSE, echo=TRUE}
# check for req'd packages, install if not present
list.of.packages <- c("tidyverse", "lubridate", "sas7bdat", "fpp2", "reshape2",
                      "stargazer", "knitcitations", "RefManageR", "xtable", 
                      "kableExtra", "zoo", "tictoc")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages, repos="http://cran.us.r-project.org")

#library loadout
library(sas7bdat)
library(fpp2)
library(zoo)
library(lubridate)
library(reshape2)
library(kableExtra)
library(knitr)
library(gridExtra)
library(tictoc)
library(tidyverse)

# set directory for lazy data referencing - allow switch between macOS and Windows
# Basically just set working directory to wherever local repo is held
setwd("~/Documents/Grad School/Thesis/github/afit.thesis/")
#setwd("C:/Users/Jake Elliott/Desktop/afit.thesis/")

# source file containing functions created for this analysis
source("~/Documents/Grad School/Thesis/github/afit.thesis/custom-functions.R")
#source("C:/Users/Jake Elliott/Desktop/afit.thesis/custom-functions.R")

```


# TO-DO

- Add complete set of lagged variable analysis
- Introduce subsetting of officer population (total vs rated-only vs non-rated line vs etc)
- Adjust legends and coloring in some figures
- Conclusion (<-- wait until lagged effects are fully explored before drawing insights)
